Wrinkling of a confined porous layer

### 2014-12-09 fig_plis/article_plis.tex
reformulations et répétitions

In the present work, we study the swelling-induced wrinkling of a porous gel sheet sandwiched between two viscous layers, with the whole system confined between two glass plates. As shown in Fig.~\ref{fig:dynamics}, the lateral confinement leads to saturation of the amplitude on the floor and ceiling followed by a cascade of secondary, ternary, etc. buckling. Such cascade buckling have been observed by~\cite{Roman1999} in a one dimensional macroscopic system, a sheet of steel forced to buckle. The 2D nature of our experiment creates nested patterns that could prove useful to design optical elements. In addition, one has to note that the primary wavelength is much smaller than the size of the system, implying that we are not in a simple buckling scenario but in a wrinkling instability. Our main result is to identify this selection mechanism.

@@

The slow acidification of a protein (sodium caseinate) solution up to the isoelectric pH induces gelation. Since caseins adsorb strongly to interfaces, the resulting gel sticks strongly to container walls, allowing rheological measurements. As shown in Fig.~\ref{fig:acidification}, further decrease of the pH beyond the isoeletric point weakens the gel. This non monotonous behaviour of the elastic modulus translates in a non monotonous mesh size in a microscopy cell (slit geometry) coated so that the casein could not adsorb on the top wall (Fig.~\ref{fig:acidification}c), i.e. in syn\ae{}resis (solvent expulsion) followed by swelling.

@@

Conversely, in a the simple case where the sheet of gel is in the middle of the cell ($H_1=H_2$) and impervious ($\alpha\rightarrow 0$), destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the lateral load across the gel $q$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim q \frac{H^3}{\eta\lambda^2}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling of the wrinkles of an elastic sheet on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural moduli of the film and substrate. Indeed, in lubrication or low Reynolds number one can consider a viscous film as elastic with effective moduli $\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

>>>>

In the present work, we study the swelling-induced wrinkling of a porous gel sheet sandwiched between two viscous layers, with the whole system confined between two glass plates. As shown in Fig.~\ref{fig:dynamics}, the lateral confinement leads to saturation of the amplitude on the floor and ceiling followed by a cascade of secondary, ternary, etc. buckling. Such cascade buckling have been observed by~\cite{Roman1999} in a one dimensional macroscopic system, a sheet of steel forced to buckle. The 2D nature of our experiment creates nested patterns that could prove useful to design optical elements. In addition, one has to note that the primary wavelength is much smaller than the size of the system, implying that we are not in a simple buckling scenario but in a wrinkling instability. Our main goal is to identify the selection mechanism of the wavelength.

@@

The slow acidification of a protein (sodium caseinate) solution up to the isoelectric pH induces gelation. Since caseins adsorb strongly to interfaces, the resulting gel sticks to container walls, allowing rheological measurements. As shown in Fig.~\ref{fig:acidification}, further decrease of the pH beyond the isoeletric point weakens the gel. This non monotonous behaviour of the elastic modulus translates in a non monotonous mesh size in a microscopy cell (slit geometry) coated so that the casein could not adsorb on the top wall (Fig.~\ref{fig:acidification}c), i.e. in syn\ae{}resis (solvent expulsion) followed by swelling.

@@

The opposite limiting case is when the sheet of gel is in the middle of the cell ($H_1=H_2$) and impervious ($\alpha\rightarrow 0$), destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the lateral load across the gel $q$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim q \frac{H^3}{\eta\lambda^2}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling of the wrinkles of an elastic sheet on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural moduli of the film and substrate. Indeed, in lubrication or low Reynolds number one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.
